---
title: "STA 244 -- Vehicle Price Prediction: A Regression Model"
subtitle: "AF Regression Project 2026"
author: "Group [XX]"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: true
    fig_height: 2.6
    fig_width: 6
geometry: margin=2cm
fontsize: 10pt
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "95%")
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2); library(ggpubr); library(scales)
  library(car); library(nortest); library(leaps)
})
theme_set(theme_bw(base_size = 8))
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))   # prediction accuracy metric
```

# Introduction and data preparation

An online marketplace wants a tool that recommends an asking `price` (in ZAR) from a
vehicle's characteristics. We build an interpretable, predictive multiple linear
regression model on `r nrow(read.csv("Assignment Information and Data sets-20260907/project_traindata.txt"))`
advertised vehicles and evaluate it out-of-sample with the **root mean square error (RMSE)**,
the metric used in the competition.

```{r load}
train_raw <- read.csv("Assignment Information and Data sets-20260907/project_traindata.txt",
                      stringsAsFactors = TRUE)
# listing_id is an identifier (not a characteristic); model_year is a perfect linear
# duplicate of vehicle_age (model_year + vehicle_age = 2026), so we keep vehicle_age only.
dat <- train_raw %>% select(-listing_id, -model_year)
```

The data contain no missing values. `price` is strongly right-skewed (mean
R`r format(round(mean(dat$price)), big.mark=",")` > median
R`r format(round(median(dat$price)), big.mark=",")`) and is bounded below by zero, so a
**log transformation of the response** is a natural candidate: it should symmetrise the
distribution, stabilise the variance, linearise the (curved) relationships with age and
mileage, and guarantee positive fitted prices.

```{r eda-response, echo=FALSE, fig.height=2.2, fig.cap="Left: price is right-skewed. Right: log(price) is approximately symmetric and its relationship with vehicle age is close to linear."}
p1 <- ggplot(dat, aes(price)) + geom_histogram(bins = 40, fill = "steelblue") +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = "price (R)", y = "count", title = "Advertised price")
p2 <- ggplot(dat, aes(vehicle_age, log(price))) +
  geom_point(alpha = 0.15, size = 0.5) + geom_smooth(se = FALSE, linewidth = 0.6) +
  labs(x = "vehicle age (years)", y = "log(price)", title = "log(price) vs age")
ggarrange(p1, p2, ncol = 2)
```

```{r eda-cats, echo=FALSE, fig.height=2.3, fig.cap="log(price) differs markedly across vehicle segments and worsens with poorer service history and reported accidents, confirming these are informative predictors."}
b1 <- ggplot(dat, aes(reorder(segment, log(price), median), log(price))) +
  geom_boxplot(fill = "grey85", outlier.size = 0.4) + coord_flip() +
  labs(x = NULL, y = "log(price)", title = "By segment")
b2 <- ggplot(dat, aes(service_history, log(price))) +
  geom_boxplot(fill = "grey85", outlier.size = 0.4) +
  labs(x = NULL, y = NULL, title = "By service history")
b3 <- ggplot(dat, aes(accident_history, log(price))) +
  geom_boxplot(fill = "grey85", outlier.size = 0.4) +
  labs(x = NULL, y = NULL, title = "By accident")
ggarrange(b1, b2, b3, ncol = 3, widths = c(1.3, 1, 0.8))
```

# Training and validation sets

To guard against over-fitting we randomly partition the data into an **80% training** set
(used to estimate every candidate model) and a **20% validation** set (used only to compare
out-of-sample prediction accuracy). An 80/20 split is standard practice: it retains enough
data for stable estimation of the many factor levels while leaving a validation sample large
enough for a reliable RMSE. A fixed seed makes the split reproducible.

```{r split}
set.seed(244)
idx   <- sample(seq_len(nrow(dat)), size = floor(0.80 * nrow(dat)))
train <- dat[idx, ]      # n = 1695
valid <- dat[-idx, ]     # n = 424
c(train = nrow(train), validation = nrow(valid), predictors = ncol(dat) - 1)
```

# Model building

**Strategy.** We work on the log scale, start from the additive model with all candidate
predictors, use significance testing and stepwise selection to remove uninformative
variables (reducing data-collection cost), then add the non-linear and interaction terms
suggested by the exploratory analysis. Each candidate is judged by validation RMSE on the
original rand scale (back-transformed), supported by adjusted $R^2$.

**Variable screening.** In the full additive log-model, `drop1` $F$-tests show every
predictor is highly significant *except* `colour` and `warranty_months`
($p \approx 0.51, 0.55$). Backward/stepwise selection by AIC agrees. We therefore drop both:
the marketplace can stop collecting exterior colour and warranty length without harming
predictive accuracy, saving data-collection and storage cost.

```{r screen, results="hide"}
full_log <- lm(log(price) ~ ., data = train)
drop1(full_log, test = "F")                 # colour & warranty_months: not significant
step_aic <- step(full_log, direction = "both", trace = 0)   # confirms same reduced set
train <- train %>% select(-colour, -warranty_months)
valid <- valid %>% select(-colour, -warranty_months)
```

**Non-linearity and interactions.** The EDA curves for age and mileage bend slightly, so we
allow second-order (quadratic) terms for both. The brief notes that a predictor's effect may
differ by segment; depreciation with age is indeed steeper for luxury vehicles than bakkies,
so we add a `segment` $\times$ age interaction. We compare four nested candidates:

```{r candidates}
f0 <- log(price) ~ .                                                        # reduced additive
f1 <- log(price) ~ . - vehicle_age + poly(vehicle_age, 2)                   # + age^2
f2 <- log(price) ~ . - vehicle_age - mileage +
        poly(vehicle_age, 2) + poly(mileage, 2)                             # + mileage^2
f3 <- log(price) ~ . - vehicle_age - mileage +
        poly(vehicle_age, 2) + poly(mileage, 2) + segment:poly(vehicle_age, 2)  # + segment x age
mods <- list(Additive = f0, `+age^2` = f1, `+mileage^2` = f2, `+seg:age` = f3)
res <- sapply(mods, function(f) {
  m  <- lm(f, data = train)
  s2 <- summary(m)$sigma^2                       # log-normal back-transform bias correction
  c(adjR2   = summary(m)$adj.r.squared,
    trRMSE  = rmse(train$price, exp(predict(m, train)) * exp(s2 / 2)),
    valRMSE = rmse(valid$price, exp(predict(m, valid)) * exp(s2 / 2)))
})
tab <- t(res)
data.frame(adjR2 = round(tab[, "adjR2"], 3),
           trainRMSE = round(tab[, "trRMSE"]),
           validRMSE = round(tab[, "valRMSE"]))
```

The `segment`$\times$age interaction gives the best validation RMSE, and its training and
validation RMSE are close, indicating the model is **not over-fitting**. This is our final
model. Because predicting on the log scale and exponentiating estimates the *median* price,
we apply the standard log-normal correction factor $\exp(\hat\sigma^2/2)$ so that fitted
values target the *mean* rand price, which lowers RMSE.

```{r final-fit}
final <- lm(f3, data = train)
sigma2 <- summary(final)$sigma^2
```

# Model diagnostics and validation of assumptions

```{r diag, echo=FALSE, fig.height=2.4, fig.cap="Residuals vs fitted show constant spread around zero with no systematic pattern (linearity + homoscedasticity); the normal Q-Q plot is close to the reference line apart from mild heavy tails."}
fit_df <- data.frame(fitted = fitted(final), resid = rstandard(final))
d1 <- ggplot(fit_df, aes(fitted, resid)) + geom_point(alpha = 0.15, size = 0.5) +
  geom_hline(yintercept = 0, colour = "red") + geom_smooth(se = FALSE, linewidth = 0.5) +
  labs(x = "fitted log(price)", y = "std. residual", title = "Residuals vs fitted")
d2 <- ggplot(fit_df, aes(sample = resid)) + stat_qq(alpha = 0.2, size = 0.5) + stat_qq_line(colour = "red") +
  labs(x = "theoretical", y = "sample", title = "Normal Q-Q")
ggarrange(d1, d2, ncol = 2)
```

The **linearity** and **constant-variance** assumptions are satisfied: residuals scatter
randomly around zero with roughly constant spread. The Q-Q plot supports approximate
**normality** with slightly heavy tails. With `r nrow(train)` observations the estimators and
$t$/$F$ inference are robust to this mild departure (CLT). **Multicollinearity** is not a
concern: generalised VIFs on the additive model are all low (below).

```{r vif}
round(sort(vif(lm(log(price) ~ ., data = train))[, 3]^2), 2)  # (GVIF^(1/2df))^2 per predictor
```

```{r influence, echo=FALSE, fig.height=1.9, fig.cap="Cook's distance: a few advertisements are mildly influential but none exceeds the 4/n rule-of-thumb by a large margin, so no observations are removed."}
infl <- data.frame(obs = seq_len(nrow(train)), cook = cooks.distance(final))
ggplot(infl, aes(obs, cook)) + geom_segment(aes(xend = obs, yend = 0), linewidth = 0.2) +
  geom_hline(yintercept = 4 / nrow(train), colour = "red", linetype = 2) +
  labs(x = "observation", y = "Cook's distance", title = NULL)
```

We inspected the highest-leverage points (very low-mileage or very high-priced luxury
listings). They are plausible genuine adverts, not data errors, so — following the brief —
**no observations are removed**.

# Interpretation and business use

Coefficients are read on the log scale, so $100(e^{\beta}-1)\%$ is the percentage change in
price for a one-unit increase in a predictor, holding others fixed.

```{r interp, echo=FALSE}
co <- summary(final)$coefficients
show <- c("transmissionManual","fuel_typePetrol","engine_size",
          "service_historyNone","service_historyPartial","accident_historyYes",
          "previous_owners","seller_typePrivate")
pct <- function(b) sprintf("%+.1f%%", 100 * (exp(b) - 1))
tab <- data.frame(term = show,
                  `pct change in price` = sapply(show, function(t) pct(co[t, 1])),
                  check.names = FALSE)
knitr::kable(tab, row.names = FALSE)
```

Key drivers: **age and mileage** dominate — price falls steeply as a car gets older and is
driven more (the quadratic terms capture depreciation flattening out for older vehicles).
A **reported accident** and **no service history** each cut price substantially, while a
**larger engine**, **automatic transmission** and **dealership** sales command a premium.
Segment strongly shifts the price level, and the `segment`$\times$age interaction shows
luxury vehicles depreciate fastest.

**How the business uses this.** For a new advert the tool returns a recommended price
$\hat{\text{price}} = \exp(\hat{\eta})\,\exp(\hat\sigma^2/2)$. Adverts whose asking price sits
far above (below) this prediction can be flagged to buyers as relatively expensive (a
bargain). The screening step also tells the marketplace it can safely stop collecting
`colour` and `warranty_months`.

# Final model and competition predictions

We refit the chosen specification on the **full training data** (train + validation) to use
all information for the final predictions, then predict the test set and write the
competition file.

```{r predict}
final_full <- lm(f3, data = dat %>% select(-colour, -warranty_months))
s2_full    <- summary(final_full)$sigma^2

test <- read.csv("Assignment Information and Data sets-20260907/project_testdata.txt",
                 stringsAsFactors = TRUE)
pred_price <- exp(predict(final_full, test)) * exp(s2_full / 2)

submission <- data.frame(listing_id = test$listing_id, price = round(pred_price))
write.csv(submission, "competition_predictions.csv", row.names = FALSE)
head(submission, 3)
```

The validated out-of-sample RMSE of approximately
R`r format(round(res["valRMSE","+seg:age"]), big.mark=",")` (against a median price of
R`r format(round(median(dat$price)), big.mark=",")`) indicates the model predicts advertised
prices well while remaining interpretable and parsimonious.
