---
title: "Predicting Used-Car Prices: A Regression Model"
subtitle: "STA 244 -- AF Regression Project 2026"
author: "Rupert van Vuuren"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: true
    fig_height: 2.5
    fig_width: 6
geometry: margin=2cm
fontsize: 10pt
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", out.width = "92%")
suppressPackageStartupMessages({
  library(dplyr); library(ggplot2); library(ggpubr); library(scales)
  library(car); library(nortest)
})
theme_set(theme_bw(base_size = 8))
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
```

# The problem and the data

The marketplace wants to suggest a sensible asking price when a seller lists a car, and to
flag adverts that look over- or under-priced. I treat this as a prediction problem and build
a multiple linear regression model for `price` (in rand) from the vehicle's characteristics.
Because the model will ultimately be judged on how well it predicts *unseen* cars, I use the
**root mean square error (RMSE)** on a held-out set as my main yardstick, alongside the
adjusted $R^2$ for in-sample fit.

```{r load}
cars <- read.csv("Assignment Information and Data sets-20260907/project_traindata.txt",
                 stringsAsFactors = TRUE)
# listing_id is just an advert number, and model_year is a perfect mirror of vehicle_age
# (model_year + vehicle_age = 2026), so keeping both would be redundant collinearity.
cars <- cars %>% select(-listing_id, -model_year)
```

There are `r nrow(cars)` cars and no missing values. A quick look at the response shows the
usual price pattern: most cars are cheap-to-mid range with a long tail of expensive ones, so
`price` is heavily right-skewed. Its relationships with the two obvious numeric drivers,
age and mileage, are also curved rather than straight — a fresh car loses value quickly and
then depreciation slows down.

```{r eda1, echo=FALSE, fig.height=2.1, fig.cap="Price is right-skewed with a long luxury tail, and both age and mileage bend the price down at a decreasing rate."}
g1 <- ggplot(cars, aes(price)) + geom_histogram(bins = 40, fill = "steelblue") +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = "price (R)", y = NULL, title = "Distribution of price")
g2 <- ggplot(cars, aes(vehicle_age, price)) + geom_point(alpha = 0.12, size = 0.5) +
  geom_smooth(se = FALSE, linewidth = 0.6) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = "age (years)", y = NULL, title = "Price vs age")
g3 <- ggplot(cars, aes(mileage, price)) + geom_point(alpha = 0.12, size = 0.5) +
  geom_smooth(se = FALSE, linewidth = 0.6) +
  scale_x_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = "mileage (km)", y = NULL, title = "Price vs mileage")
ggarrange(g1, g2, g3, ncol = 3)
```

Among the categorical variables, `segment` and `service_history` stand out as clearly
price-related, and a reported accident visibly lowers the typical price. `make`, `fuel_type`,
`transmission` and `seller_type` show smaller but real differences, while `colour` and
`province` look almost flat against price — something I return to when selecting variables.

```{r eda2, echo=FALSE, fig.height=2.2, fig.cap="Price varies strongly by segment and service history and drops when an accident is reported; colour shows little relationship."}
c1 <- ggplot(cars, aes(reorder(segment, price, median), price)) +
  geom_boxplot(fill = "grey85", outlier.size = 0.3) + coord_flip() +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = NULL, title = "By segment")
c2 <- ggplot(cars, aes(service_history, price)) + geom_boxplot(fill = "grey85", outlier.size = 0.3) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = NULL, title = "By service history")
c3 <- ggplot(cars, aes(colour, price)) + geom_boxplot(fill = "grey85", outlier.size = 0.3) +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = NULL, title = "By colour") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggarrange(c1, c2, c3, ncol = 3, widths = c(1.2, 1, 1.3))
```

# Training and validation split

To get an honest read on prediction accuracy I set aside part of the data before doing any
modelling. I use an 80/20 split — 80% to estimate models and 20% held back purely to compare
them — which keeps enough data to estimate the many factor levels reliably while leaving a
validation set large enough to give a stable RMSE. I fix the seed so the split (and every
result that follows) is reproducible.

```{r split}
set.seed(244)
i     <- sample(nrow(cars), floor(0.80 * nrow(cars)))
train <- cars[i, ]
valid <- cars[-i, ]
c(train = nrow(train), validation = nrow(valid))
```

# Building the model

**Transforming the response.** The skew and the curved relationships above both point to
modelling `log(price)` rather than `price`. To check this properly rather than by eye, I run
a Box-Cox analysis on the full model. The estimated power is $\hat\lambda \approx 0.06$ with a
confidence interval covering 0, so the log ($\lambda = 0$) is the transformation the data
support. It also has three practical pay-offs: it straightens out the age and mileage curves,
makes the error spread roughly constant, and guarantees the model never predicts a negative
price.

```{r boxcox, results="hide", fig.show="hide"}
bc <- boxCox(lm(price ~ ., data = train), plotit = FALSE)
bc$x[which.max(bc$y)]   # ~ 0.06  ->  log transform
```

**Choosing variables.** Starting from a model with every predictor, both `drop1` $F$-tests
and stepwise selection by AIC agree that `colour` and `warranty_months` add nothing
($p \approx 0.5$) — exactly what the flat colour boxplot suggested. Everything else is
significant. I drop these two, which also gives the business a concrete saving: it can stop
recording exterior colour and warranty length without hurting the price estimates.

```{r select, results="hide"}
full <- lm(log(price) ~ ., data = train)
drop1(full, test = "F")               # colour & warranty_months not significant
step(full, direction = "both", trace = 0)  # AIC keeps the same reduced set
train <- train %>% select(-colour, -warranty_months)
valid <- valid %>% select(-colour, -warranty_months)
```

I did consider collapsing rare categories into an "Other" group, but every level of `make`
(smallest 81 cars) and `province` is well populated and jointly significant, so grouping would
only throw away information.

**Adding shape and interactions.** From the exploration I let age and mileage enter as
quadratics to capture the flattening depreciation, and — since the brief hints that a
predictor's effect can differ by segment — I allow the age effect to vary by segment (luxury
cars lose value faster than bakkies). I compare each addition on the validation set, using the
log-normal correction $\exp(\hat\sigma^2/2)$ when back-transforming so the predictions target
the mean rand price rather than the median.

```{r compare}
forms <- list(
  "additive"        = log(price) ~ .,
  "+ age^2"         = log(price) ~ . - vehicle_age + poly(vehicle_age, 2),
  "+ age^2, mile^2" = log(price) ~ . - vehicle_age - mileage +
                        poly(vehicle_age, 2) + poly(mileage, 2),
  "+ segment:age"   = log(price) ~ . - vehicle_age - mileage +
                        poly(vehicle_age, 2) + poly(mileage, 2) + segment:poly(vehicle_age, 2)
)
compare <- sapply(forms, function(f) {
  m  <- lm(f, data = train); s2 <- summary(m)$sigma^2
  c(adjR2     = summary(m)$adj.r.squared,
    trainRMSE = rmse(train$price, exp(predict(m, train)) * exp(s2 / 2)),
    validRMSE = rmse(valid$price, exp(predict(m, valid)) * exp(s2 / 2)))
})
data.frame(adjR2 = round(compare["adjR2", ], 3),
           trainRMSE = round(compare["trainRMSE", ]),
           validRMSE = round(compare["validRMSE", ]))
```

Each step lowers the validation RMSE, and the segment-by-age interaction gives the best result
without pulling the training and validation errors apart, so it is not just fitting noise. I
tried several further interactions (segment with mileage, engine size and transmission, and a
`make`$\times$`segment` term); none improved the validation RMSE and the `make`$\times$`segment`
version clearly over-fit, so I stopped here.

```{r final}
final  <- lm(forms[["+ segment:age"]], data = train)
sigma2 <- summary(final)$sigma^2
```

# Checking the assumptions

```{r diag, echo=FALSE, fig.height=2.3, fig.cap="Residuals scatter evenly around zero with roughly constant spread, and the Q-Q plot follows the line closely apart from slightly heavy tails."}
rd <- data.frame(fit = fitted(final), r = rstandard(final))
p1 <- ggplot(rd, aes(fit, r)) + geom_point(alpha = 0.12, size = 0.5) +
  geom_hline(yintercept = 0, colour = "red") + geom_smooth(se = FALSE, linewidth = 0.5) +
  labs(x = "fitted log(price)", y = "standardised residual", title = "Residuals vs fitted")
p2 <- ggplot(rd, aes(sample = r)) + stat_qq(alpha = 0.15, size = 0.5) + stat_qq_line(colour = "red") +
  labs(x = "theoretical quantiles", y = "sample quantiles", title = "Normal Q-Q")
ggarrange(p1, p2, ncol = 2)
```

The residual plot shows no leftover curve and a fairly even band, so the linearity and
constant-variance assumptions hold up well on the log scale. The Q-Q plot is close to straight
with mildly heavy tails; a formal Anderson-Darling test does reject perfect normality
($p \approx 0.003$), but with `r nrow(train)` observations the coefficient estimates and their
tests are reliable regardless (central limit theorem), so I am comfortable with this. There is
no multicollinearity problem either — every variance-inflation factor on the additive model is
below about 4.5.

```{r infl, echo=FALSE, fig.height=1.8, fig.cap="A handful of high-priced luxury cars are mildly influential, but none dominates the fit, so nothing is removed."}
ci <- data.frame(obs = seq_len(nrow(train)), cook = cooks.distance(final))
ggplot(ci, aes(obs, cook)) + geom_segment(aes(xend = obs, yend = 0), linewidth = 0.2) +
  geom_hline(yintercept = 4 / nrow(train), colour = "red", linetype = 2) +
  labs(x = "observation", y = "Cook's distance", title = NULL)
```

The most influential points are expensive, low-mileage luxury cars. I checked them
individually; they are plausible genuine adverts, not data-entry errors, so I keep them — the
model should be able to price such cars, and removing them just to flatter the fit would be
dishonest.

# What the model says, and how the business uses it

Because the response is logged, each coefficient $\beta$ translates into a
$100(e^{\beta}-1)\%$ change in price for a one-unit change in that variable, holding the rest
fixed. The main drivers are exactly the ones a buyer would expect:

```{r interp, echo=FALSE}
co <- summary(final)$coefficients
terms <- c("service_historyNone", "accident_historyYes", "engine_size",
           "transmissionManual", "seller_typePrivate", "previous_owners")
labs  <- c("No service history (vs full)", "Accident reported (vs none)",
           "Engine size (+1 litre)", "Manual (vs automatic)",
           "Private seller (vs dealership)", "Each extra previous owner")
knitr::kable(data.frame(Factor = labs,
             `Effect on price` = sprintf("%+.1f%%", 100 * (exp(co[terms, 1]) - 1)),
             check.names = FALSE))
```

Age and mileage dominate the price (their quadratic terms trace the steep-then-flattening
depreciation curve), and `segment` shifts the whole price level — a hatchback sells for around
44% less than a comparable bakkie, while luxury cars sit highest. On top of that, condition
and history matter: a car with no service record loses about 16% and a reported accident about
11%, so honest, well-serviced cars command a clear premium. Bigger engines and dealership sales
lift the price, and each extra previous owner shaves a little off.

For the marketplace this gives a ready-made pricing tool: for a new advert it returns a
recommended price $\hat{\text{price}} = \exp(\hat\eta)\,\exp(\hat\sigma^2/2)$, and any advert
sitting far above or below that number can be flagged to buyers as expensive or a bargain. The
model also tells the business which fields are worth collecting — age, mileage, segment, make,
service and accident history carry almost all the signal, while colour and warranty length can
be dropped to save data-collection cost. Its main weakness is the luxury tail: those cars carry
the largest rand errors simply because small percentage misses translate into large amounts, so
I would treat luxury price suggestions as more approximate.

# Final predictions

For the competition I refit the chosen model on all `r nrow(cars)` cars (training plus
validation) so the final predictions use every bit of data, then predict the test set.

```{r predict}
final_all <- lm(forms[["+ segment:age"]], data = cars %>% select(-colour, -warranty_months))
s2_all    <- summary(final_all)$sigma^2

test <- read.csv("Assignment Information and Data sets-20260907/project_testdata.txt",
                 stringsAsFactors = TRUE)
pred <- exp(predict(final_all, test)) * exp(s2_all / 2)
write.csv(data.frame(listing_id = test$listing_id, price = round(pred)),
          "competition_predictions.csv", row.names = FALSE)
```

On the held-out validation set this model predicts price to within an RMSE of about
R`r format(round(compare["validRMSE", "+ segment:age"]), big.mark=",")`, against a typical
price near R`r format(round(median(cars$price)), big.mark=",")` — accurate enough to be useful
while staying simple enough to explain to a seller.
